\newcommand{\cost}{\text{cost}}
\newcommand{\car}{\text{car}}
\newcommand{\pt}{\text{PT}}
\newcommand{\walk}{\text{walk}}
\newcommand{\bike}{\text{bike}}
%\newcommand{\dummy}[1]{\delta_{#1}}
\newcommand{\hi}{\text{h.i}}
\newcommand{\wait}{\text{wait}}
\newcommand{\geqfive}{\text{s.z.,walk}}
\newcommand{\mc}{\theta}
\newcommand{\ac}{C}
\newcommand{\acp}{\theta}
\newcommand{\dura}[1]{\Delta t_#1}
\newcommand{\TT}{TT}
\newcommand{\ustay}{\avgu_\text{stay}}
\newcommand{\stay}{\text{stay}}

\newcommand{\p}[1]{p_{\text{#1}}}
\newcommand{\m}[1]{m_{\text{#1}}}
\newcommand{\amemm}[1]{\amem_\text{#1}}
\newcommand{\dummy}[1]{\delta_{#1}}

This section consist of a case study where a model is specified and estimated based on the method proposed in Section \ref{seq:mdp}. 

We have limited this case study to the special case when there is no uncertainty in the transition of the state variables in $x_k$ and when $\epsilon$ is i.i.d Gumbel distributed. As discussed in \ref{sec:estSampling}, this special case can be estimated using sampling of alternative sequences of actions. Limiting the case study to this special case has the benefit that the parameter estimates obtained represent the 'true' model, i.e., a model where no biasing approximations are made. These parameter values can thus be used to evaluate the bias introduced by the approximations of the value function.
We use the estimates to generate statistics of, e.g., the departure time of trips and the number of trips used. We also simulate aggregate travel times, travel costs etc., to compare the model output with the data. 

% Description of what type of behaviour we set out to model in our case study.
We model individuals who work with fairly regular working hours on a day when they go to work. We assume that work is a mandatory activity as well as dropping of children at school or picking them up after school in case such activities are observed. People can have fixed working schedule, in which case the starting time for job is fixed, or flexible, in which case we assume that they are allowed to arrive between $6\unit{a.m.}$ and $10\unit{a.m.}$.
Children can be dropped off between $6:30\unit{a.m.}$ and $12:00\unit{a.m.}$ and picked up between $12\unit{a.m.}$ and $6:30\unit{p.m.}$
In the case study, an individual is defined by an observed sequence of actions $\ve{a}^n$ together with the following attributes:
\begin{equation}
i^n = 
\begin{pmatrix}
l^n_\text{home} \\
l^n_\text{work} \\
l^n_\text{child} \\
\dummy{car} \\
\dummy{pt card} \\
\dummy{flex} \\
t^n_\text{work}\\
\\
P^n_{\text{mandatory}}
\\
\text {}
\end{pmatrix} = 
\begin{pmatrix*}[l]
\text{home location} \\
\text{work location} \\
\text{child location} \\
\text{car ownership dummy} \\
\text{pt card dummy} \\
\text{flexible working hours dummy} \\
\text{work duration} \\
\\
\text{list of mandatory activities}
\\
\text {}
\end{pmatrix*}
\in
\begin{Bmatrix*}[l]
\{1,2,\dots, N_L\} \\
\{1,2,\dots, N_L\} \\
\{1,2,\dots, N_L\} \\
\{0,1\} \\
\{0,1\} \\
\{0,1\} \\
[0,T] \\
\begin{Bmatrix*} [\p{work}] ,[\p{child},\p{work}] \\  
[\p{work},\p{child}] 
\\ [\p{child},\p{work},\p{child}]\end{Bmatrix*}
\end{Bmatrix*}
\end{equation}
\subsection{Model specification}
Below we discuss how the MDP is specified in the case study. This involves defining the actions, states, conditional choice sets, state transitions and one-stage utility functions.  
\subsubsection{Actions}
Remember that an action consisted of the choice of a location, activity and mode. We have used the so called EMME zonal system of the region of Stockholm, which is standard to use in models in the area and for which official statics are available. The Stockholm region consist of 1240 such zones, so $N_L = 1240$. The modes modelled are car, public transport, bike and walk. Activities are divided into home, work, child (pick up or drop off), social, recreational, grocery shopping and other.

\begin{equation} 
a_k = \begin{pmatrix}
\act{d} \\
\act{m} \\
\act{p}    
\end{pmatrix}
\in
\begin{Bmatrix*}[l]
\{1,2,\dots, 1240\}, \\
\{m_{\text{stay}},\m{car},\m{pt},\m{walk},\m{bike}\}, \\
\{ p_{\text{travel}},p_{\text{end}},\p{home},\p{work},\p{child},\p{rec.},\p{social},\p{shop},\p{other}\} 
\end{Bmatrix*}
\end{equation}
\subsubsection{States}
We let a day start at $5\unit{am}$ and end at $11\unit{pm}$, and let $t$ define the number of $10\units{minute}$ intervals since $5\unit{am}$ so that $T=108$.
We will for the case study let the marginal utility of activity participation be constant and use $\tau$ only to remember if the activity has been started ($\tau=1$) or not ($\tau=0$), so $N_\atime=\max{N_{\atime,p}}=1$. The mode history will keep track of whether car is used as the mode of transport on a tour. The mode history can thus be described by three values: $m_{\text{stay}},\m{car}$ and $\m{pt}$. The activity history stores the number of completed activities in the list of mandatory activities where the number of mandatory activities for an individual $n$ is $N^n_\text{mandatory}\leq 3$. 
\begin{equation}
\s{k} = \begin{pmatrix}
t \\
l \\
p \\
\atime\\
m \\
\amem \\
\epsilon
\end{pmatrix}
\in
\begin{Bmatrix*}[l]
[0,108] \\
\{1,2,\dots, 1240\} \\
\{ \p{home},\p{work},\p{child},\p{rec.},\p{social},\p{shop},\p{other}\}  \\
\{0\} \\
\{m_{\text{stay}},\m{car},\m{pt} \}\\
\{0,1,2,3\}\\
\rr^{N_C}
\end{Bmatrix*}
\end{equation}

\subsubsection{Conditional choice-set}
The choice set $C^n(s_k)$ for an individual $n$ requires the specification of $M^n(m)$, $L^n_{\text{act}}(\act{p})$ and $P^n(h)$.
The available modes are:
\begin{equation*}
M^n(m) = \begin{cases}
\{ \m{pt},\m{walk},\m{bike} \} & \cif \dummy{car}^n = 0 \\
\{ \m{car},\m{pt},\m{walk},\m{bike} \} & \cif \dummy{car}^n = 1,m = \m{stay} \\
\{ \m{car}\} & \cif m = \m{car} \\
\{ \m{pt},\m{walk},\m{bike} \} & \cif m = \m{pt} 
\end{cases}
\end{equation*}
Let $P_{\text{freetime}} = \{\p{rec.},\p{social},\p{shop},\p{other}\}$. The locations available for a specific individual $n$ is:
\begin{equation}
L^n_{\text{act}}(\act{p}) = \begin{cases}
L & \cif \act{p} \in P_{\text{freetime}} \\
l^n_\text{home} & \cif \act{p} = \p{home} \\
l^n_\text{work} & \cif \act{p} = \p{work} \\
l^n_\text{child} & \cif \act{p} = \p{child} \\
\end{cases}
\end{equation}
Remember that for each individual we have an ordered list of mandatory activities $P^n_{\text{mandatory}}$ where $P^n_{\text{mandatory}}(i)$ is a set containing the $i$:th activity in list. The available activities conditional on history can now be defined as:
\begin{equation}
P^n(h) = P_{\text{freetime}} \cup \{\p{home}\} \cup \begin{cases}
P^n_{\text{mandatory}}(h+1)& \cif h < N^n_\text{mandatory} \\
\emptyset & \celse
\end{cases}
\end{equation}
\subsubsection{State transitions}
The unspecified state transition are $q_t$, $q_h$ and $f_m$. As $m$ is used to remember whether car was the first trip on a tour away from home it is defined as:
\begin{equation}
    m\ks{k+1} = 
    \begin{cases}
        \m{car} & \cif \act{m} = \m{car} \text{ and } m=m_\text{stay} \\
        \m{pt}  & \cif \act{m} \neq \m{car} \text{ and } m=m_\text{stay}  \\
        m_\text{stay} & \cif \act{p} = p_{\text{home}} \\
		m\ks{k} & \celse
    \end{cases}
\end{equation}
With $\amem$ denoting the number of finished mandatory activities, it is updated whenever a 
\begin{equation}
	\amem\ks{k+1} = 
	\begin{cases}
		\amem\ks{k}+1 &\cif \act{p} = P^n_{\text{mandatory}}(\amem\ks{k}) \\
		\amem\ks{k}& \celse.
	\end{cases}
\end{equation}
Finally, when there is no travel time uncertainty, $t\ks{k+1}$ is simply given by:
\begin{equation}
	t\ks{k+1} =t\ks{k} + \begin{cases}
  	\TT_{\act{m}}(t\ks{k},l\ks{k},\act{d}) & \cif m\neq m_\text{stay} \\
  	t_{\text{min}}^n(\act{p}) & \cif m= m_\text{stay} \text{, } \atime = 0  \\
  	\Delta_p t & \cif m= m_\text{stay} \text{, } \atime \geq 1 
	\end{cases}
\end{equation}

\subsubsection{One-stage utility functions}

The one-stage utility of a state-action pair was in \eqref{eq:mdponestage} dividable into four parts: the (dis)utility of travelling $\utravel(t,l,m,\act{d},\act{m})$; the utility from activity participation $\uact(t,p,\atime)$; the utility representing preferences for when to start and where to perform an activity $\ustart(l,t,\amem, p)$; and a utility component for interaction between current and future purpose $\uchange(p,\act{p})$ which is left out from this case study. Below follows the specification of these two utility functions used in the case study, starting with the utility of travelling.
\paragraph{Utility of travelling}
The utility of travelling is given by:
\begin{equation} \label{eq:utravel}
u_{trav}(t,l,m,\act{d},\act{m}) = \mc_{\act{m}} + \theta_{tt,\act{m}} \cdot \TT_{\act{m}}(t,l,\act{d}) \theta_{\cost} \cdot C_{\act{m}}(t,l,\act{d}) +\theta_{\geqfive} \dummy{\geqfive}
\end{equation}
where $\mc_{\act{m}}$ are mode specific constants; $\theta_{tt,\act{m}}$ are mode specific (dis)utilities for travel time  $\TT_{\act{m}}(t,l,\act{d})$ and $C_{\act{m}}(t,l,\act{d})$ denote the travel time and cost with mode $\act{m}$ at time $t$ for a trip from origin $l$ to destination $\act{d}$; and $\dummy{\geqfive}$ is a dummy indicating if the trip is done within the same zone with walk. Observe that the mode state $m$ does not influence the one-stage utility in the case-study specification.

\paragraph{Utility of starting activity}
The utility of starting an activity $\ustart(l,t,\amem, p)$ is here independent of $\amem$ and given by:
\begin{equation}
	\ustart(l,t,p) = \begin{cases}
	u_{\text{size}}(l,p) & \cif p \in P_{\text{freetime}} \\
	u_\text{child constraints}(t) & \cif p = \p{child} \\
	u_\text{work constraints}(t)  + u_{\text{start work}}(t) & \cif p = \p{work} \\
	0 & \cif p = \p{home}  
	\end{cases} 
\end{equation}
where $u_{\text{size}}$ models the number of opportunities that exist for each purpose in a specific location; $u_\text{constraints}$ models constraints on when activities can be performed; and $u_{\text{start time}}$ models preferences for when to start specific activities. The size utility is given by:
\begin{equation*}
u_{p,\text{size}}(l) =\theta_{C,p} + \theta_{p,\text{size}}\log \left( \sum_{s=1}^{s=S_p} x_{p,l,s}e^{\theta_{p,s}} \right)
\end{equation*}
where $S_p$ is the number of size variables for activity $p$, and the size variables $x_{p,l,s}$ can be, e.g., the number of employees in a specific sector at location $l$. Since the size utility contain an activity specific constant $\theta_{C,p}$, one of the parameters $\theta_{p,s}$ should be fixed for all activities. This also provides an alternative interpretation of the activity specific constants $\theta_{C,p}$ as scales for the size variables $x_{p,l,s}$. A complete list of size variables included for respectively activity is given in table \ref{tab:ustart}. 

Child and work constraints are ensure that these activities are started within the opening hours of day-care and schools in the case of child activities and within the acceptable starting hours for work for the individual, so:
\begin{equation*}
\begin{aligned}
		u_\text{child constraints}(t) &= 
		\begin{cases}
				0 & \cif t_{\text{drop off after}} \leq t \leq t_{\text{drop off before}} \\
		0 & \cif t_{\text{pick up after}} \leq t \leq t_{\text{pick up before}} \\
		-\infty & \celse
		\end{cases} 
		\\
		u_\text{work constraints}(t) &= 
		\begin{cases}
		0 & \cif t_{\text{work start after}}^n \leq t \leq t_{\text{work start before}}^n \\
		-\infty & \celse
		\end{cases}
\end{aligned}
\end{equation*}

The time-of-day dependent utility of starting work is specified at a number of points in time $t_j$ by parameters $\acp_{\text{work},t_j}$. If work starts at a time $t$ between two such time point $t_j$ and $t_{j+1}$, i.e., if $t_j\leq t \leq t_{j+1}$, linear interpolation is used to obtain the start utility:
\begin{equation*}
u_{\text{start work}}(t)  = \frac{\acp_{\text{work},j}\cdot (t_{j+1}-t)+\acp_{\text{work},j+1}\cdot (t-t_j)}{t_{j+1}-t_{j}}
\end{equation*}
\paragraph{Utility of activity participation}
The utility of activity participation is modelled as:
\begin{equation} \label{eq:uact}
	\uact(t,p,\atime) = \begin{cases} 
	\theta_{t,p} \cdot \Delta_t &\cif p \in P_{\text{freetime}} \\
	u_\text{stay home}(t) & \cif p = p_\text{home} \\
	0 & \celse
	\end{cases}
\end{equation}
\newcommand{\home}{\text{home}}
 For the utility of staying at home for another time step, we specify the \emph{marginal} utility of activity participation at time $t$ as given by linearly interpolation between the closest parameters, so:
\begin{equation*}
u_\text{marginal stay home}(t) = \frac{\theta_{\home,t_j}\cdot(t_{j+1}-t_j)+\theta_{\home,t_{k+1}}\cdot(t-t_j) }{t_{j+1}-t_{j}}.
\end{equation*}
The stay-utility of an activity episode of duration $\dura{p}$, when $T_k\leq t$ and $t+\dura{p}\leq T_{k+1}$, then becomes:
\begin{equation} \label{eq:uacttime}
\begin{aligned}
u_\text{stay home}(t) &= \int_t^{t+\dura{p}} u_\text{marginal stay home}(y)
\ud{y} = \alpha_{t_j}\theta_{\home,t_j} + \alpha_{t_{j+1}}\theta_{\home,t_{j+1}}% \\
%&= \frac{\theta_{p,T_k}T_{k+1}-\theta_{p,T_{k+1}}T_k + (\theta_{p,T_{k+1}} - \theta_{p,T_{k}}) (2t\dura{p} + \dura{p}^2)}{T_{k+1}-T_{k}}.
\end{aligned}
\end{equation}
where:
\begin{align*}
\alpha_{t_j}&=\dura{p}\frac{t_{j+1} - t - 0.5 \dura{p}}{t_{j+1}-t_{j}}\\
\alpha_{t_{j+1}}&=\dura{p}\frac{t + 0.5\dura{p} - t_j}{t_{j+1}-t_{j}}.
\end{align*}
If $t+\dura{p}> t_{j+1}$, $u_\text{marginal stay home}(y)$ in \refeq{eq:uacttime} becomes a stepwise linear function and additional factors $\alpha_{t_{j+2}}$ is added to the utility function.

\subsubsection{Computational details}
\label{seq:computationTime}

With the current implementation, calculating the value function in all states and thus evaluating the probability of a path once for a single individual takes between $4-10\units{s}$. Almost all ($98\%$) of the computation time is consumed by the function calculating the value function in all end-activity states when summing up the alternative trips that can be started. When excluding working hours, an example individual have 11\,h of free time left between 5:00AM and 23:00PM. The value function is evaluated on a 10\,minute grid, giving 65 grid points in which it will be evaluated. In each of these grid points, there are a total of 4 modes available for a car owner and 1\zdel 240 locations that are both states and destinations. This gives a total of $65 \cdot 4 \cdot 1\zdel 240 \cdot 1\zdel 240 \sim 4 \cdot 10^8$ links. For each of these links, one must calculate the travel time (taking $16\%$ of the time), one-stage utilities ($23\%$), obtain the future expected value function and sum this with the utility ($41\%$) and finally perform the exponent $e^{u_i + EV_j}$ for all links ($20\%$). All-in-all, this takes around 4-10\,s when using a single core on a Intel(R) Core(TM) i7-6820HQ CPU @ 2.70GHZ. The main program is written in C\# but Intel MKL has been used for vector mathematics when applicable and C++ routines has been used for other time consuming parts. As a comparison, simply performing $e^x$ in MATLAB for a vector $x$ with $10^8$ elements on the same computer takes 1\,s. 

With $10\units{s/observation}$ to calculate $\eutil$, $70$ variables, $3\zdel 300$ observations and $100$ iterations before convergence, this would require $10 \cdot 0.4 \cdot  70 \cdot 3\zdel 300 \cdot 100 \units{s} \approx 1\zdel 000 \units{days}$ to estimate using a single core.
The program is parallelized using MPI and could potentially benefit from hundreds of cores reducing the computation time to days. A possible future solution to obtain efficient estimates and allow for nested logit or GEV formulations would be to use inefficient estimates obtained using sampling of locations or some alternative approximate or possibly biased estimation technique to find a good enough specification of the model and then finally obtain full-information estimates using a cluster. 

 As discussed above, the from-all-to-all destinations operation is currently the limiting computational factor. One possible way to speed up the model evaluation would be to sample locations. If 100 locations were sampled, the computation time could potentially decrease to $0.05\units{s/individual}$.  


\subsection{Data}
\input{text/caseStudy/Data.tex}

\subsection{Estimation result}
%TODO Discuss how estimation is done

\input{text/caseStudy/EstResult.tex}

\subsection{Simulation result}
\input{text/caseStudy/Simulation.tex}


